---
title: "Lab4: Clustering (and PCA)"
author: "STAT 4610/5610"
format: pdf
editor: visual
date: 2026-04-14
---

## **Due: 11:59PM Friday, April 17 on Canvas as a knitted pdf file of your team's Quarto document**

1.  You will individually create clusters from the `Attitude` data set that some of you have contributed to. You will create clusters of the data points (the respondents) AND also clusters of the attitudes.
2.  You will individually interpret the clusters of the data points and compare them to the established `Group` of respondents.
3.  As a team, you will decide the best way to cluster this data set. You will interpret these clusters.

### The Context

Six groups of individuals responded to a survey about Attitudes in Collaboration conducted by Prof. Vance and Laily Mualifah, Dr. Vance's colleague in Indonesia (where he spent the 2023-24 academic year as a Fulbright Scholar experimenting how to teach interdisciplinary collaboration to Indonesian statistics and data science students). The goal of the research project is to see what differences in attitudes are held by Indonesian students compared to Indonesian professionals compared to students in the U.S.

If you haven't taken the survey, you still can (for 5 bonus points) at https://forms.gle/Q7d5XVMeub5Y2DLN6 until 7:59AM Tuesday, April 14. By Tuesday at noon, the dataset will be finalized and released. You can take the survey later, but not for bonus points. (Note: Don't take the survey a second time.)

#### The Survey

The survey asks respondents to rate how often they have an attitude and then to what degree they think that attitude is detrimental to or beneficial for interdisciplinary collaboration. Then some demographic and experience questions are asked at the end of the survey.

The six groups who took the survey were:

-   Indonesian professionals (not undergraduate students) in 2024, labeled `IndoProf24`

-   Indonesian undergraduates in 2024 after taking a course on Statistical Collaboration, labeled `IndoPost24`

-   Indonesian undergraduates in 2025 at the beginning of taking a course on Statistical Collaboration, labeled `IndoPre25`

-   Students from Statistical Collaboration at CU Boulder. Both ugrads and grad students, labeled `StatCollab`

-   Students from Statistical Learning at CU Boulder in Spring 2025 labeled `SLSp25` and Spring 2026 labeled `SLSp26`.

#### The Data

The data can be downloaded from: <https://drive.google.com/file/d/179nQH4lO2oUR_9l8sySFKZI7RDPU5rMG/view?usp=sharing> (this is new data to be updated at 11:30AM on April 14.)

The questions were asked in groups of five. The first five questions were about how often the respondent had Attitudes 1-5, then how beneficial/detrimental Attitudes 1-5 would be for collaboration, then how often for Attitudes 6-10 and how beneficial/detrimental for Attitudes 6-10, and so on until Attitudes 21-25.

The column name "A1.freq" indicates Frequency of Attitude 1 (How often do you have this attitude?) on a scale from 1 (Never) to 5 (Always). The ratings numbers 1-5 correspond to the ratings "Never" "Rarely" "Sometimes" "Usually" and "Always" (and their Indonesian translations) for the Frequency questions.

The column name "A1.fav" indicates how detrimental/beneficial the respondent thinks that attitude is on a six-point scale: "Extremely Detrimental (1)" "Detrimental (2)" "Somewhat Detrimental (3)", "Somewhat Beneficial (4)", "Beneficial (5)", and "Extremely Beneficial (6)" (and their Indonesian translations).

The 25 attitudes are downloadable here: <https://docs.google.com/document/d/18Hjt4LPC2vkko9Bxm_7NqVgKYLEto8Mz/edit?usp=sharing&ouid=108624562691464780001&rtpof=true&sd=true>

The `Group` variable in the data set has values "StatCollab", "SLSp25", "SLSp26", "IndoProf24", "IndoPost24", and "IndoPre25" corresponding to American students in Statistical Collaboration F24/Sp25, students in Statistical Learning Sp25 and Sp26, Indonesian professionals in 2024, Indonesian undergraduates in 2024 after taking a course on Statistical Collaboration, and Indonesian undergraduates in 2025 before taking a course on Statistical Collaboration.

### Instructions for Lab4

#### What individuals need to do (20 pts)

```{r}
library(tidyverse)
library(cluster)
library(factoextra)
library(dendextend)

survey <- Attitude_clean

groups <- survey$Group

X <- survey %>%
  select(-Group)

X_scaled <- scale(X)
```

1.  Implement K-means clustering on the data points. Use a sufficient number of starting values (20 is probably fine). Choose $K\in \{2,3,4,5,6\}$. Each individual should choose a different $K$.

```{r}
set.seed(1056)

k6 <- kmeans(X_scaled, centers = 6, nstart = 20)

respondent_clusters <- k6$cluster

table(respondent_clusters, groups)
```

1.  Interpret your results. To what degree do your $K$ clusters correspond to the four groups?

```{r}
pr.out <- prcomp(X_scaled, scale = TRUE)

plot_df <- data.frame(
  PC1 = pr.out$x[,1],
  PC2 = pr.out$x[,2],
  Cluster = as.factor(k6$cluster),
  Group = survey$Group
)

ggplot(plot_df, aes(PC1, PC2, color = Cluster, shape = Group)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_minimal()
```

The K = 6 K means clustering matched the original groups somewhat, but not very clearly. A few clusters were made up mostly of one group, like StatCollab, which suggests those respondents answered in similar ways. However, most of the other groups were spread across several clusters, especially SLSp25 and SLSp26, showing a lot of overlap in responses. The Indonesian groups also appeared together in some clusters rather than separating cleanly. Overall, the clustering found some patterns in the data, but the six clusters did not strongly correspond to the original groups, suggesting that attitudes toward collaboration were fairly mixed across the different populations.

1.  For the same $K$, cluster the attitudes (cluster the transpose of the $X$ matrix). Interpret your results. Comment on the patterns you see within the clusters of attitudes.

```{r}
X_t <- t(X_scaled)

k6_att <- kmeans(X_t, centers = 6, nstart = 20)

attitude_clusters <- data.frame(
  Attitude = rownames(X_t),
  Cluster = k6_att$cluster
)

attitude_clusters %>% arrange(Cluster)
```

The respondents tended to rate how often they experienced an attitude similarly to how beneficial they believed that attitude was. Some clusters, like Cluster 2 and Cluster 6, contained larger groups of attitudes, which may represent broader themes in how respondents view collaboration related behaviors, while smaller clusters like Cluster 3 and Cluster 4 contained only one attitude pair, suggesting those attitudes were more unique in their response patterns. Overall, the results show that certain attitudes were consistently perceived in similar ways, indicating that respondents may group related collaboration attitudes together rather than viewing each one independently.

1.  Use hierarchical clustering to cluster the data points. Each individual should use a different linkage (complete, single, average, or centroid). You must try at least two different distance measures (e.g., Euclidean, Manhattan, correlation). Comment on how many clusters you chose and why? Comment on which distance measure you used for your final clusters and why.

```{r}
dist_euc <- dist(X_scaled, method = "euclidean")
hc_euc <- hclust(dist_euc, method = "complete")
plot(hc_euc, labels = FALSE, main = "Complete Linkage - Euclidean")
```

```{r}
dist_man <- dist(X_scaled, method = "manhattan")
hc_man <- hclust(dist_man, method = "complete")
plot(hc_man, labels = FALSE, main = "Complete Linkage - Manhattan")
```

```{r}
hc_clusters <- cutree(hc_euc, k = 6)

table(hc_clusters, groups)

hcman_clusters <- cutree(hc_man, k = 6)

table(hcman_clusters, groups)
```

For hierarchical clustering, I used complete linkage and compared Euclidean and Manhattan distance measures, choosing 6 clusters so the results could be compared directly to the six original survey groups. Between the two methods, Euclidean distance gave a more balanced clustering structure, while Manhattan distance placed most respondents into one very large cluster, making the results less informative. Even with Euclidean distance, the clusters only partially matched the original groups. The largest cluster contained respondents from every group, especially StatCollab, IndoPre25, and IndoPost24, which suggests that many respondents across groups shared similar collaboration attitudes. Some smaller clusters showed mild concentrations of IndoProf24 respondents, but no group formed a clearly separate cluster on its own. Overall, the hierarchical clustering showed that while there are some differences among groups, the response patterns overlap substantially, and the original survey groups are not strongly separated based on these attitudes.

1.  Plot the "best" dendrogram. Interpret your results. To what degree do your $K$ clusters correspond to the four groups?

```{r}
dist_euc <- dist(X_scaled, method = "euclidean")

hc_euc <- hclust(dist_euc, method = "complete")

plot(hc_euc,
     labels = FALSE,
     hang = -1,
     main = "Hierarchical Clustering Dendrogram",
     xlab = "",
     sub = "")
```

The hierarchical clustering with complete linkage and Euclidean distance produced only a weak to moderate correspondence with the original survey groups. After cutting the dendrogram into k=6 clusters, one large cluster contained respondents from all groups, especially StatCollab, IndoPre25, and IndoPost24, suggesting these groups share similar response patterns rather than forming distinct clusters. The remaining clusters were smaller and more mixed, with no group forming a clean, isolated cluster. Overall, while the clustering identified some structure in the data, the K clusters do not closely match into groups.

1.  Cluster the attitudes using hierarchical clustering. Use a different linkage (or use more than one and pick the best one). Justify your choice of distance measure.

```{r}
dist_att <- dist(t(X_scaled), method = "euclidean")
hc_att <- hclust(dist_att, method = "average")

plot(hc_att, main = "Hierarchical Clustering of Attitudes")
rect.hclust(hc_att, k = 6, border = "red")

att_clusters <- cutree(hc_att, k = 6)

data.frame(
  Attitude = colnames(X_scaled),
  Cluster = att_clusters
) %>% arrange(Cluster)
```

I used hierarchical clustering with average linkage and Euclidean distance to cluster the attitudes. Euclidean distance was appropriate because the variables are on a comparable scale and captures overall similarity in response patterns. Average linkage was chosen because it provides a balance between overly loose and overly strict clustering, producing more interpretable groups.

The results show clear structure in the attitudes. One large cluster contains many paired frequency and favorability items, suggesting groups of attitudes that respondents tend to evaluate similarly. Smaller clusters isolate a few distinct attitudes that behave differently from the rest. Overall, the attitudes form several coherent groups rather than acting independently, indicating underlying patterns in how respondents perceive collaboration related behaviors.

1.  Plot the "best" dendrogram. Interpret your results. Compare the patterns you see within the hierarchical clusters of attitudes with the patterns you saw using K-means.

```{r}
dist_att <- dist(t(X_scaled), method = "euclidean")
hc_att <- hclust(dist_att, method = "average")

plot(hc_att,
     main = "Hierarchical Clustering of Attitudes",
     xlab = "",
     sub = "",
     cex = 0.6,
     hang = -1)
```

The hierarchical clustering and K means results show similar groupings of attitudes, especially where frequency and favorability versions of the same items cluster together. However, hierarchical clustering provides more detail by showing how attitudes gradually merge into larger groups, while K means forces a fixed partition. Overall, both methods identify consistent patterns, but hierarchical clustering is more interpretable because it reveals the structure within the attitude relationships rather than just final group assignments.

#### What teams need to do

1.  Review and combine individual results. Which methods for clustering the data points seemed to correspond best with the pre-established groups? What $K$ gave you the best results? What linkage? What distance metric?

1.  Compare the results from Principal Component Analysis (PCA) with the results of clustering the attitudes. Which is more interpretable? What is your interpretation?

1.  Create your team's final, best clustering of the data points. How well do the clusters correspond to the four groups?

1.  Create your team's final, best clustering of the attitudes. What patterns do you see within the clusters of attitudes?

1.  Write a paragraph about individual contributions. Each individual must comment on their contributions to the lab. Only individuals who contribute to the team section will get points for the team section. An example contribution could be: "I synthesized all of the individual results to find that such and such gave the best and most interpretable results. Then I recommended teammate X use this linkage and that distance measure for our team's cluster." Or, "I did the comparison of PCA with our clustering of the attitudes." Or, "I created our team's best clustering of the data points based on teammate Y's synthesis of our individual results." Or, "I took everyone's code and put it together into the qmd file and knitted it." Or, "I made the plot of our team's best clusters."

Ryan: Compiled each individual’s contributions into one final document and prepared the submission. In addition, Ryan completed team sections 3 through 5, including creating the team’s final clustering of the data points, interpreting how well those clusters corresponded to the pre established groups, and summarizing the patterns observed in the final clustering of the attitudes.

Emery: Completed team section 2 by comparing the PCA results with the clustering of the attitudes, evaluating which method was more interpretable, and providing an interpretation of the differences between the two approaches.

Jake: Completed team section 1 by reviewing the individual clustering results from the team, comparing the methods used for clustering the data points, and determining which values of K, linkage methods, and distance metrics gave the best correspondence to the pre established groups.

#### Some intended outcomes from this assignment:

1.  You will individually get practice using K-means and hierarchical clustering
2.  You will practice using different distance measures and linkages for hierarchical clustering
3.  You will gain experience interpreting the outputs of clustering algorithms
4.  You will gain experience collaborating with your teammates on an applied problem

#### Potentially useful code below

#### 

```{r include=F}
suppressPackageStartupMessages(library(tidyverse))
dat <- read_csv("Attitude.clean.csv")
dim(dat)

X <- dat[,-1] #Excluding the groups
dim(X)

# Creating distance measures
cor_spearman <- cor(t(X), method = "spearman") # Transpose so the rows are the attitude variables
dist_spearman <- as.dist(1 - cor_spearman)

cor_pearson <- cor(t(X), method = "pearson") 
dist_pearson <- as.dist(1 - cor_pearson)

# Using Manhattan or Euclidean distance
dist_manhattan <- dist(X, method = "manhattan") #Manhattan distance counts 
# the "blocks" between data points as if you were walking along the streets of 
# Manhattan (i.e., you can't cut diagonally through a city block)
dist_euclidean <- dist(X, method = "euclidean") #Euclidean distance is the 
# diagonal "shortest path" between points A and B.

# Using hclust
hc.complete <- hclust(dist_spearman, method = "complete")
cutree(hc.complete, k=4)

# Kmeans
km.out <- kmeans(X, k=4, nstart = 20)

# Doing PCA
pr.out <- prcomp(X,scale=F) #The attitudes are basically already on the same scale, so scaling the attitudes would confuse things, though A1.freq is on a 5-point scale and A1.fav is on a 6-point scale

# Find means of the groups, complete for the other two groups
SL25.means <- colMeans(dat[dat[,1]=="SLSp25",2:51])
StatCollab.means <- colMeans(dat[dat[,1]=="StatCollab",2:51])
IndoProf.means <- colMeans(dat[dat[,1]=="IndoProf24",2:51])
IndoPost.means <- colMeans(dat[dat[,1]=="IndoPost24",2:51])

# Then transform the group centroids into the Principal Component space
C.matrix <- rbind(StatCollab.means,SL25.means,IndoProf.means,IndoPost.means)

#Transform C.matrix into scores using the rotation matrix of the first two principal components
W2 <- pr.out$rotation[,1:2]
Group.scores <- C.matrix %*% W2
Group.scores <- mutate(as.data.frame(Group.scores),groups=rownames(Group.scores))
ggplot(data=Group.scores, aes(x=PC1,y=PC2,col=groups)) +
  geom_point()

# Comparing found clusters with groups
# What percentage of each cluster is each group?
table(km.out$cluster,dat$Group) #can transform this into percentages if necessary

# We can do the same with clusters found from hiearchical clustering
table(cutree(hc.complete,k=4),dat$Group) #These clusters aren't very good.

```
